---
title: 'Applied Regression: Homework 4'
author: "Group 7"
date: "December 4, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(AER)
```

```{r}
# import data
setwd("D:/Econometrics/Program Evaluation/HW/Applied-Regrassion-Analysis/HW4")
acs = read.table("Cal.csv", header=T, sep=',', stringsAsFactors = F)

# Data cleaning
acs$uhrswork[acs$uhrswork=='99 (Topcode)'] = 99
acs$uhrswork = ifelse(acs$uhrswork=='N/A', 0, acs$uhrswork)
acs$uhrswork = as.numeric(acs$uhrswork)
acs$workedyr = ifelse(acs$workedyr=='Yes', 1, 0)
```
# Q1: Verify that the same sex variable has predicts having more children with and without the covariates.
```{r}
model.wo = lm(morekids ~ samesex, data = acs)
summary(model.wo)

model.wi = lm(morekids ~ samesex + as.factor(age) + white + black + asian + hisp, data = acs)
summary(model.wi)
```
without covariates: 0.038**
with covariates: 0.041**
# Q2: Using OLS, depict the impact of having more children on each of the three outcome variables.
```{r}

model.2.1 = lm(incearn ~ morekids + as.factor(age) + white + black + asian + hisp, data = acs)
model.2.2 = lm(uhrswork ~ morekids + as.factor(age) + white + black + asian + hisp, data = acs)
model.2.3 = lm(workedyr ~ morekids + as.factor(age) + white + black + asian + hisp, data = acs)

summary(model.2.1)
summary(model.2.2)
summary(model.2.3)

```
incearn: -11173.5***
uhrswork: -4.8531***
workedyr: -0.13139***
#Q3: Using two-stage least squares, depict the impact of having more children on each of the three outcome variables. Use the variable for “samesex” variable as the instrument.
```{r}
# First Stage
acs$morekids_pred = fitted.values(model.wi)

# Second Stage
model.3.1 = lm(incearn ~ morekids_pred + as.factor(age) + white + black + asian + hisp, data = acs)
model.3.2 = lm(uhrswork ~ morekids_pred + as.factor(age) + white + black + asian + hisp, data = acs)
model.3.3 = lm(workedyr ~ morekids_pred + as.factor(age) + white + black + asian + hisp, data = acs)

summary(model.3.1)
summary(model.3.2)
summary(model.3.3)

# IVReg
model.3.4 = ivreg(incearn ~ morekids + as.factor(age) + white + black + asian + hisp | samesex + as.factor(age) + white 
                  + black + asian + hisp, data = acs)
model.3.5 = ivreg(uhrswork ~ morekids + as.factor(age) + white + black + asian + hisp | samesex + as.factor(age) + white 
                  + black + asian + hisp, data = acs)
model.3.6 = ivreg(workedyr ~ morekids + as.factor(age) + white + black + asian + hisp | samesex + as.factor(age) + white 
                  + black + asian + hisp, data = acs)

summary(model.3.4)
summary(model.3.5)
summary(model.3.6)
```
incearn: 6238.0
uhrswork: 4.33114
workedyr: -0.11090
# Q4: Using the test given in BJLST, test for selection bias for each of the three outcomes. Be sure to estimate the selection bias.
```{r}
# BJLST test for selection bias

# test on Y1
model.4.1 = lm(incearn ~ samesex + as.factor(age) + white + black + asian + hisp, data = acs[acs$morekids==1,])
model.4.2 = lm(uhrswork ~ samesex + as.factor(age) + white + black + asian + hisp, data = acs[acs$morekids==1,])
model.4.3 = lm(workedyr ~ samesex + as.factor(age) + white + black + asian + hisp, data = acs[acs$morekids==1,])

summary(model.4.1)
summary(model.4.2)
summary(model.4.3)

# test on Y0
model.4.4 = lm(incearn ~ samesex + as.factor(age) + white + black + asian + hisp, data = acs[acs$morekids==0,])
model.4.5 = lm(uhrswork ~ samesex + as.factor(age) + white + black + asian + hisp, data = acs[acs$morekids==0,])
model.4.6 = lm(workedyr ~ samesex + as.factor(age) + white + black + asian + hisp, data = acs[acs$morekids==0,])

summary(model.4.4)
summary(model.4.5)
summary(model.4.6)

```
Y1:
incearn -544.1
uhrswork  0.1026
workedyr -0.009324

Y0:
incearn 1306.5
uhrswork  0.5449
workedyr  0.007447
